Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-01-05 analysis/paper_draft3.Rmd
added to discussion

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity). Current statistical approaches to the detection of bias have two extreme approaches to inclusion of historical data regardless of incompatibility between current and historical trials. Firstly,  historical data can be ignored entirely. This approach may lead to underpowered estimates of bias due to the low number of strikes in a single trial (typically 6, 10, or 15 strikes per party), and may lead to inflated estimates of bias. Secondly, historical and current trial data can be combined and analysed together. While this approach increases the sample size, it ignores the possible effects of incompatiblity between current and historical trial.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. [editme] The goal of these methods is not to _prove_ bias, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge.
The Bayesian approach here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can easily assess how sensitive that posterior estimate is to the available historical strike data. [editme] sentence on why this is appealing

Moreover, with a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias inference is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to inferring bias based on historical strike data has that ability.

>>>>

In turn, the inference of bias based in part on historical strike data is relevant to lawyers for deciding whether to argue a _Batson_ violation, proving a prima facie case, and ultimately persuading the trial judge that bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)). One challenge for incorporating historical data is possible incompatibility between current and historical strike data for a given attorney that may arise because of missing data or because attorney bias depends on certain trial-level characteristics (e.g., defendant race, charge severity). Current statistical approaches to the detection of bias have two extreme approaches to inclusion of historical data regardless of incompatibility between current and historical trials. Firstly,  historical data can be ignored entirely. This approach may lead to underpowered estimates of bias due to the low number of strikes in a single trial (typically 6, 10, or 15 strikes per party), and may lead to inflated estimates of bias. Secondly, historical and current trial data can be combined and analysed together. While this approach increases the sample size, it ignores the possible effects of incompatiblity between current and historical trial. Such incompatibility can lead to inflated type I error and reduced power [@Bennett2020].

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. [editme] The goal of these methods is not to _prove_ bias, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge. Previously used methods have either ignored historical trial data, or conducted analyses where current and historical data are merged, with no regard for possible incompatibility between trials. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not.

Moreover, with a tool like our prototype app, attorneys can use the approach presented here to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias inference is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to inferring bias based on historical strike data has that ability. In future work we hope to examine data-driven approached to the selection of $\alpha$; where the degree of incompatibility can be estimated and used to inform the choice of the degree of discounting to apply to historical data. [editme - refs?] Knowing how to quantify incompatibility and how to specify and appropriate prior $\alpha$ is an area of active research [e.g., https://arxiv.org/abs/2206.04379]
